---
title: "Figures_v3"
output: html_document
date: "2025-11-26"
---

**Figures for the manuscript titled, "Ecological, evolutionary, and social drivers of Oropuche virus expansion within and beyond the Amazon"** 


Figure questions can be directed to Samantha Sambado (ssambado at stanford.edu)


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## upload necessary packages
library(cowplot)
library(ggpubr)
library(ggplot2)
library(ggspatial)
library(janitor)
library(paletteer)
library(patchwork)
library(rgbif)
library(rnaturalearth)
library(rnaturalearthdata)
library(scales)
library(sf)
library(terra) 
library(tidyterra)
library(tidyverse)


```

#Spatial Boundaries
```{r}
### baseline map for figures
# whole world
world <- ne_countries(scale = "large", returnclass = "sf")

## central/south america
cs_america <- world %>% 
  filter(geounit %in% c("Belize", "Costa Rica", "El Salvador", "Guatemala", 
                        "Honduras", "Nicaragua", "Panama", "Argentina", "Bolivia", 
                        "Brazil", "Chile", "Colombia", "Ecuador", "Guyana",
                        "Paraguay", "Peru", "Suriname", "Uruguay", "Venezuela", "French Guiana",
                        "Barbados", "Cuba", "Dominica", "Haiti", "Trinidad and Tobago"))

# make vect for mask/crop
cs_america_vect <- vect(cs_america)


### administrative boundaries
admin <- c("Belize", "Costa Rica", "El Salvador", "Guatemala", "Honduras", "Nicaragua", "Panama",
           "Argentina", "Bolivia", "Brazil", "Chile", "Colombia", "Ecuador", "Guyana",
           "Paraguay", "Peru", "Suriname", "Uruguay", "Venezuela", "French Guiana",
           "Barbados", "Cuba", "Dominica", "Haiti", "Trinidad and Tobago")

admin_countries <- ne_states(country = admin, returnclass = "sf")

```

#Figure 2 - Human OROV
```{r}
## data
# manually scraped data from PAHO OROV portal
# https://www.paho.org/en/arbo-portal/arbo-portal-oropouche
orov_cases <- read.csv("data/raw/OROV_PAHO_samentered.csv") %>% 
  mutate(country = str_to_title(country), # fix country labeling for merging
    country = case_when(country == "Brasil" ~ "Brazil",
                        TRUE ~ country))

## calculate average incidence from PAHO OROV portal
cs_america_inc <- cs_america %>% 
  left_join(orov_cases, by = c("admin" = "country")) %>% 
  # laboratory confirmed incidence (2024/2025 or just single reported year)
  mutate(incidence = case_when(admin == "Barbados" ~ (0.66),
                               admin == "Bolivia" ~ (3.01),
                               admin == "Brazil" ~ (6.47+5.60)/2,
                               admin == "Colombia" ~ (0.16+0.05)/2,
                               admin == "Cuba" ~ (5.53+0.31)/2,
                               admin == "Ecuador" ~ (0.02),
                               admin == "Guyana" ~ (0.38+0.13)/2,
                               admin == "Panama" ~ (0.36+14.74)/2,
                               admin == "Peru" ~ (3.97+0.99)/2,
                               admin == "Venezuela" ~ (0.02),
                               admin == "Chile" ~ 0,
                               admin == "Uruguay" ~ 0,
                               TRUE ~ NA)) %>%
  mutate(incidence = ifelse(is.na(incidence), 0, incidence)) 

## make map labels
cs_centroids <- cs_america_inc %>%
  st_point_on_surface() %>%
  mutate(lon = st_coordinates(.)[,1],
         lat = st_coordinates(.)[,2],
         label = iso_a2) %>% 
  # get rid of labels that take up too much space compared to country
  filter(iso_a2 != "BZ" &
           iso_a2 != "GT" &
           iso_a2 != "HN" &
           iso_a2 != "NI" &
           iso_a2 != "CR" &
           iso_a2 != "SV" &
           iso_a2 != "DM" &
           iso_a2 != "HT" &
           iso_a2 != "TT" &
           iso_a2 != "PA" &
           iso_a2 != "BB" &
           iso_a2 != "CU" &
           iso_a2 != "SR" &
           iso_a2 != "GY" &
           iso_a2 != "EC")

## plot it!

## map of average incidence per country
map_incidence <- ggplot() +
  geom_sf(data = cs_america_inc, aes(fill = incidence), color = "grey20", size = 0.3) +
  coord_sf(expand = FALSE) + 
  scale_fill_gradient(low = "grey90", high = "grey10", 
                      na.value = "white", name = "Average OROV Incidence") +
  geom_sf(data = admin_countries, fill = NA, color = alpha("grey10", 0.25)) +
  geom_sf_label(data = cs_centroids, aes(label = label), 
                size = 3, fill = "white",color = "black",fontface = "bold", label.size = .1) +
  theme_void(base_size = 13) +
  theme(legend.title = element_text(face = "bold"),
        plot.margin = margin(2, 2, 2, 2), # changed from all 0s
        legend.position = "bottom",
        legend.key.height = unit(0.35, "cm"),
        legend.key.width = unit(1.2, "cm")) +
  guides(fill = guide_colorbar(title.position = "top")) +
  annotation_scale(width_hint = 0.3, location = "br") + 
  annotation_north_arrow(which_north = "true",
                         location = "br", pad_y = unit(1, "cm"),
                       style = north_arrow_fancy_orienteering)

## barplot of weekly cases
barplot_cases <- orov_cases %>% 
  ggplot(aes(x = week, y = case, fill = country)) +
  geom_bar(width = 0.85, color = "grey20", linewidth = 0.2,stat = "identity") +
  theme_classic(base_size = 13) +
  labs(x = "Epidemiological Week", y = "Human OROV Cases", fill = "") +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_fill_paletteer_d("Redmonder::dPBIRdGn", direction = -1) +
  theme(legend.title = element_text(face = "bold"),
        axis.title = element_text(face = "bold"),
        legend.position  ="bottom",#"right", #
        legend.box.spacing = unit(0,"pt"),
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(.1, "lines"),
        strip.background = element_rect(fill = "grey90", color = NA),
        strip.text = element_text(face = "bold", size = 13)) +
  facet_wrap(~year) + 
  guides(fill = guide_legend(nrow = 2, byrow = TRUE))


### combine them
combo_orovcases <- map_incidence | barplot_cases +
                plot_layout(widths = c(2.5, 1)) 

#ggsave(file ="figures/combo_orovcases.png", combo_orovcases,width = 12, height = 5.5, units = "in", dpi = 300)
```

#Figure 3 - Environmental & Social Drivers of EIDs

###Part A - temperature

wrangle data
```{r}
### step 1. upload CHELSA data
# set params
years  <- 2015:2020  # change if needed
months <- sprintf("%02d", 1:12) # change if needed

# call CHELSA website
base <- "https://os.unil.cloud.switch.ch/chelsa02/chelsa/global/monthly/tas"

urls <- c()

for (y in years) {
  for (m in months) {
    urls <- c(urls, sprintf("%s/%d/CHELSA_tas_%s_%d_V.2.1.tif", base, y, m, y))}
}

# may need to optimize for computer
options(timeout = 1000)
for (url in urls) {
  outfile <- file.path("CHELSA_tas", basename(url))
  download.file(url, outfile, mode = "wb", method = "curl")
}

### step 2. crop CHELSA to south america
# need to figure out better optimization for my computer
terraOptions(memfrac = 0.6, tempdir = "temp/")

# call files in 
chelsa_files <- list.files("data/raw/CHELSA_tas", pattern = "\\.tif$", full.names = TRUE)
chelsa_r <- rast(chelsa_files)
cs_america_vect2 <- project(cs_america_vect, crs(chelsa_r))

# set up some basics
extent_sa <- ext(-95, -30, -60, 25)
in_dir  <- "data/raw/CHELSA_tas"
out_dir <- "data/processed/CHELSA_tas_SA"


# set up forloop to crop 
dir_files <- list.files(in_dir, pattern = "\\.tif$", full.names = TRUE)

for (f in dir_files) {
  message("Processing: ", basename(f))
  
  r <- rast(f)

  out_file <- file.path(out_dir, basename(f))

  # crop + mask & write directly to disk
  r_mask <- mask(crop(r, extent_sa),
                 cs_america_vect2,
                 filename = out_file,
                 overwrite = TRUE)

  gc()  #clear memory between iterations
}

### step 3. calculate average temperature during rainy season
in_dir2 <- "data/processed/CHELSA_tas_SA/"
crop_files <- list.files(in_dir2, pattern = "\\.tif$", full.names = TRUE)

# set params of years & months 
years <- 2015:2020
rain_months <- c("12", "01", "02", "03", "04", "05", "06", "07") 

# build pattern to pull specific params
year_pattern <- paste(years, collapse = "|")
month_pattern <- paste(rain_months, collapse = "|")
pattern <- paste0("_(", month_pattern, ")_(", year_pattern, ")_V\\.2\\.1\\.tif$")

rain_files <- crop_files[grepl(pattern, crop_files)]

# read stack & compute mean
rain_stack <- rast(rain_files)
mean_temp_K <- mean(rain_stack, na.rm = TRUE)
mean_temp_C <- mean_temp_K - 273.15 # convert to C
#plot(mean_temp_C)

out_file <- "data/processed/chelsa_meantemp_rainyseason_20152020.tif"
#writeRaster(mean_temp_C, out_file, overwrite = TRUE)
```

plot it!
```{r}

## set up data
mean_temp_C <- rast("data/processed/chelsa_meantemp_rainyseason_20152020.tif")
temp_df <- as.data.frame(mean_temp_C, xy = TRUE)
colnames(temp_df) <- c("lon", "lat", "temp")

## make plot
map_temp_rainyszn <- ggplot() +
  geom_raster(data = temp_df, aes(x = lon, y = lat, fill = temp)) +
  geom_sf(data = cs_america, fill = NA, color = alpha("grey40", 0.8), size = 0.35) +
  coord_sf(xlim = c(-95, -30), ylim = c(-60, 25), expand = FALSE) +
  theme_void(base_size = 13) +
  labs(fill = "°C", title = "Mean Temperature") +
  theme(legend.title = element_text(face = "bold"),
        plot.margin = margin(0, 0, 0, 0)) +
  scale_fill_paletteer_c("grDevices::RdYlBu", direction = -1) 

```


###Part B - forest cover

wrangle data
```{r}
## my computer hated this, need to optimize
# forest cover files
cover_files <- list.files("data/raw/hansen_forest_cover/", pattern = "cover_.*\\.tif$", full.names = TRUE)

# make virtual raster
forest_cover <- vrt(cover_files)

# reclassify forest as 30%
rcl_mat <- matrix(c(-Inf, 30, 0,
                    30, Inf, 1), ncol = 3, byrow = TRUE)

# Classify & write directly to disk (this somewhat helped)
forest_cover_bin <- writeRaster(
  classify(forest_cover, rcl = rcl_mat),
  filename = "data/processed/forest_cover_bin.tif",
  overwrite = TRUE,
  datatype = "INT1U",  # 1-byte unsigned integer is enough for 0/1
  wopt = list(datatype = "INT1U", gdal = "COMPRESS=LZW"))
  
#plot(rast("data/processed/forest_cover_bin.tif"))


```

plot it!
```{r}
## set up data
forest_cover_bindata <- rast("data/processed/forest_cover_bin.tif")

## plot it
map_forest_hansen <- ggplot() +
  # kept to spatraster to not explode my computer as geom_sf
  geom_spatraster(data = forest_cover_bindata) +
  scale_fill_gradient(low = "transparent", high ="#778868FF" ) + 
  theme_void() +
  geom_sf(data = cs_america, fill = NA, color = alpha("grey40", 0.8), size = 0.35) +
  coord_sf(xlim = c(-95, -30), ylim = c(-60, 25), expand = FALSE) +
  theme_void(base_size = 13) +
  labs(title = "Forest Cover") +
  theme(legend.title = element_text(face = "bold"),
        legend.position = "bottom",
        legend.key.width = unit(2, "cm"),
        legend.key.height = unit(0.4, "cm"),
        legend.margin = margin(0, 0, 0, 0),
        legend.box.margin = margin(0, 0, 0, 0),
        plot.margin = margin(0, 0, 0, 0)) +
  guides(fill = FALSE) +
  annotation_scale(width_hint = 0.3, location = "br") + 
  annotation_north_arrow(which_north = "true",
                         location = "br", pad_y = unit(1, "cm"),
                       style = north_arrow_fancy_orienteering)


```

###Part C - population density

wrangle data
```{r}
pop <- terra::rast("data/raw/gpw-v4-population-density-rev11_2015_30_sec_tif/gpw_v4_population_density_rev11_2015_30_sec.tif")

# buffer polygon to avoid holes
poly_buf <- buffer(cs_america_vect, width = 0.05)

# crop and mask
pop_crop <- crop(pop, poly_buf)
pop_mask <- mask(pop_crop, poly_buf, touches = TRUE)

# convert to df
pop_df <- as.data.frame(pop_mask, xy = TRUE)
colnames(pop_df) <- c("lon", "lat", "pop_density")

# avoid log10 issues 
pop_df$pop_density_adj <- pop_df$pop_density + 0.001

```

plot it!
```{r}
map_pop <- ggplot() +
  geom_raster(data = pop_df, aes(x = lon, y = lat, fill = pop_density_adj)) +
  geom_sf(data = cs_america, fill = NA, color = alpha("grey40", 0.8), size = 0.35) +
  coord_sf(xlim = c(-95, -30), ylim = c(-60, 25), expand = FALSE) +
  theme_void(base_size = 13) +
  labs(fill = "people/km²", title = "Population Density") + 
  theme(legend.title = element_text(face = "bold"),
        plot.margin = margin(0, 0, 0, 0)) +
  scale_fill_paletteer_c("ggthemes::Blue-Green Sequential",
                         trans = "log10", direction = -1, 
                         breaks = c(0.001, 0.1, 1, 10, 100, 1000, 10000),
                         labels = c(c("0", "0.1", "1", "10", "100", "1k", "10k")))

```


###Part D - healthcare access

wrangle data
```{r}
travel <- rast("data/raw/2020_walking_only_travel_time_to_healthcare.geotiff")
travel_crop <- crop(travel, ext(cs_america_vect))
travel_sa <- mask(travel_crop, cs_america_vect)

```

plot it!
```{r}
## set up data
travel_df <- as.data.frame(travel_sa, xy = TRUE, na.rm = TRUE)
colnames(travel_df) <- c("lon", "lat", "traveltime")

# convert minutes to hours
travel_df$traveltime_hr <- travel_df$traveltime / 60

map_travel <- 
ggplot() +
  geom_raster(data = travel_df, aes(x = lon, y = lat, fill = traveltime_hr)) +
  geom_sf(data = cs_america, fill = NA, color = alpha("grey40", 0.8), size = 0.35) +
  coord_sf(xlim = c(-95, -30), ylim = c(-60, 25), expand = FALSE) +
  theme_void(base_size = 13) +
  labs(fill = "hours", title = "Access to Healthcare") +
  theme(legend.title = element_text(face = "bold"),
        plot.margin = margin(0, 0, 0, 0)) +
  scale_fill_paletteer_c("grDevices::RdYlBu", direction = -1, trans = "log1p",
                         breaks = c(1, 2, 4, 8, 24, 48), 
                         labels = c("1", "2", "4", "8", "24", "48")) +
  guides(fill = guide_colorbar(barheight = 10.0))




```

#Figure 4 - Hypothesized Ecological Components

###Part A - crop types
wrangle data
```{r}
#### BANANA
banana_rast <- terra::rast("data/raw/CROPGRIDSv1.08_NC_maps/CROPGRIDSv1.08_banana.nc")
banana <- banana_rast[["harvarea"]]
banana_crop <- crop(banana, cs_america_vect)
banana_mask <- mask(banana_crop, cs_america_vect)
banana_df <- as.data.frame(banana_mask, xy = TRUE)
colnames(banana_df) <- c("lon", "lat", "harvest")


### COCOA
cocoa_rast <- terra::rast("data/raw/CROPGRIDSv1.08_NC_maps/CROPGRIDSv1.08_cocoa.nc")
cocoa <- cocoa_rast[["harvarea"]]
cocoa_crop <- crop(cocoa, cs_america_vect)
cocoa_mask <- mask(cocoa_crop, cs_america_vect)
cocoa_df <- as.data.frame(cocoa_mask, xy = TRUE)
colnames(cocoa_df) <- c("lon", "lat", "harvest")


## SOYBEAN
soybean_rast <- terra::rast("data/raw/CROPGRIDSv1.08_NC_maps/CROPGRIDSv1.08_soybean.nc")
soybean <- soybean_rast[["harvarea"]]
soybean_crop <- crop(soybean, cs_america_vect)
soybean_mask <- mask(soybean_crop, cs_america_vect)
soybean_df <- as.data.frame(soybean_mask, xy = TRUE)
colnames(soybean_df) <- c("lon", "lat", "harvest")


```

plot it!
```{r}
# banana
map_banana <- ggplot() +
  geom_raster(data = banana_df, aes(x = lon, y = lat, fill = harvest)) +
  geom_sf(data = cs_america, fill = NA, color = alpha("grey40", 0.8), size = 0.35) +
  coord_sf(xlim = c(-95, -30), ylim = c(-60, 25), expand = FALSE) +
  theme_void(base_size = 13) +
  labs(fill = "ha", title = "Banana") +
  theme(legend.title = element_text(face = "bold"),
        plot.margin = margin(0, 0, 0, 0),
        legend.position = "right",
        legend.title = element_text(face = "bold", size = 10),
        legend.text = element_text(size = 9),
        legend.key.height = unit(0.45, "cm"),
        legend.key.width  = unit(0.6, "cm"),
        legend.margin = margin(2, 2, 2, 2),
        legend.box.margin = margin(2, 2, 2, 2),
        legend.spacing = unit(0.3, "cm")) +
  guides(fill = guide_colorbar(barheight = unit(4, "cm"),
                               barwidth  = unit(0.5, "cm"),
                               ticks.colour = "black",
                               frame.colour = "black")) +
  scale_fill_paletteer_c("viridis::magma", trans = "log1p",
                         breaks = c(0, 1, 10, 100, 1000),
                         direction = -1) 

# cocoa
map_cocoa<- ggplot() +
  geom_raster(data = cocoa_df, aes(x = lon, y = lat, fill = harvest)) +
  geom_sf(data = cs_america, fill = NA, color = alpha("grey40", 0.8), size = 0.35) +
  coord_sf(xlim = c(-95, -30), ylim = c(-60, 25), expand = FALSE) +
  theme_void(base_size = 13) +
  labs(fill = "ha", title = "Cocoa") +
  theme(legend.title = element_text(face = "bold"),
        plot.margin = margin(0, 0, 0, 0),
        legend.position = "right",
        legend.title = element_text(face = "bold", size = 10),
        legend.text = element_text(size = 9),
        legend.key.height = unit(0.45, "cm"),
        legend.key.width  = unit(0.6, "cm"),
        legend.margin = margin(2, 2, 2, 2),
        legend.box.margin = margin(2, 2, 2, 2),
        legend.spacing = unit(0.3, "cm")) +
  guides(fill = guide_colorbar(barheight = unit(4, "cm"),
                               barwidth  = unit(0.5, "cm"),
                               ticks.colour = "black",
                               frame.colour = "black")) +
  scale_fill_paletteer_c("viridis::magma", trans = "log1p",
                         breaks = c(0, 1, 10, 100, 1000),
                         direction = -1) 

# soybean
map_soybean <- ggplot() +
  geom_raster(data = soybean_df, aes(x = lon, y = lat, fill = harvest)) +
  geom_sf(data = cs_america, fill = NA, color = alpha("grey40", 0.8), size = 0.35) +
  coord_sf(xlim = c(-95, -30), ylim = c(-60, 25), expand = FALSE) +
  theme_void(base_size = 13) +
  labs(fill = "ha", title = "Soybean") +
  theme(legend.title = element_text(face = "bold"),
        plot.margin = margin(0, 0, 0, 0),
        legend.position = "right",
        legend.title = element_text(face = "bold", size = 10),
        legend.text = element_text(size = 9),
        legend.key.height = unit(0.45, "cm"),
        legend.key.width  = unit(0.6, "cm"),
        legend.margin = margin(2, 2, 2, 2),
        legend.box.margin = margin(2, 2, 2, 2),
        legend.spacing = unit(0.3, "cm")) +
  guides(fill = guide_colorbar(barheight = unit(4, "cm"),
                               barwidth  = unit(0.5, "cm"),
                               ticks.colour = "black",
                               frame.colour = "black")) +
  scale_fill_paletteer_c("viridis::magma", trans = "log1p",
                         breaks = c(0, 1, 10, 100, 1000),
                         direction = -1) 


### combine them
# make common legend
common_scale <- scale_fill_paletteer_c("viridis::magma", trans = "log1p", direction = -1,
                                       breaks = c(0, 1, 10, 100, 1000),
                                       labels = c("0","1","10","100","1000"),
                                       na.value = "#E5E5E5FF", name = "ha")

# make slight adjustments
map_banana2 <- map_banana + common_scale + theme(legend.position = "none")
map_cocoa2  <- map_cocoa  + common_scale + theme(legend.position = "none")
map_soybean2 <- map_soybean + common_scale + theme(legend.position = "none")

# combine
maps_crop_combo <- (map_banana2 | map_cocoa2 | map_soybean2) + 
  plot_layout(guides = "collect") +
  theme(legend.position = "right",               
        legend.title = element_text(face = "bold", size = 10),
        legend.text  = element_text(size = 9),
        legend.key.height = unit(1.5, "cm"),    
        legend.key.width  = unit(0.4, "cm"),    
        legend.margin = margin(2,2,2,2),
        legend.box.margin = margin(2,2,2,2))

```

###Part B - host range

wrangle data
```{r}
iucn_folder <- "data/raw/iucn_range/"

# list all species folders 
species_folders <- list.dirs(iucn_folder, recursive = FALSE)
species_folders <- species_folders[grepl("redlist_species_data_", species_folders)]


read_iucn_species <- function(folder) {
  # shapefile path
  shp_file <- file.path(folder, "data_0.shp")
  
  # Read shapefile
  sp <- st_read(shp_file, quiet = TRUE)
  
  # Extract species name from folder name
  # Split by "_" and take the last element
  sp_name <- tail(strsplit(basename(folder), "_")[[1]], 1)
  
  # Add species column
  sp$species <- sp_name
  return(sp)
}

species_list <- lapply(species_folders, read_iucn_species)
species_sf <- do.call(rbind, species_list)

```

plot it!
```{r}
map_host <- ggplot() +
  geom_sf(data = species_sf, aes(fill = SCI_NAME)) +
  geom_sf(data = cs_america, fill = NA, color = alpha("grey40", 0.8), size = 0.35) +
  coord_sf(xlim = c(-95, -30), ylim = c(-60, 25), expand = FALSE) +
  theme_void(base_size = 13) +
  labs(fill = "Species", title = "Host Range") + 
  theme(legend.title = element_text(face = "bold"),
        plot.margin = margin(0, 0, 0, 0)) +
  scale_fill_paletteer_d("PNWColors::Spring")

```

###Part C - vector occurrence

wrangle 
```{r}
## gbif credentials
Sys.setenv(GBIF_USER = "sbsambado")
Sys.setenv(GBIF_PWD = "2411ucsb")
Sys.setenv(GBIF_EMAIL = "sbsambado@ucsb.edu")

# extents
xlim <- c(-95, -30)
ylim <- c(-60, 25)

# make bounding box
csa_shp_wkt <- sprintf(
  "POLYGON((%f %f, %f %f, %f %f, %f %f, %f %f))",
  xlim[1], ylim[1],  # xmin, ymin
  xlim[2], ylim[1],  # xmax, ymin
  xlim[2], ylim[2],  # xmax, ymax
  xlim[1], ylim[2],  # xmin, ymax
  xlim[1], ylim[1])   # close polygon

## call in gbif data
culi <- name_backbone(name = "Culicoides", rank = "genus")
culi_key <- culi$usageKey

gbif_culi <- occ_download(
  pred_and(
    pred("geometry", csa_shp_wkt),
    pred("hasCoordinate", TRUE),
    pred("taxonKey", culi_key),
    pred("occurrenceStatus", "PRESENT")),
  format = "SIMPLE_CSV")

# check download status
occ_download_wait('0002605-251120083545085')


# work with gbif data
gbif_data <- occ_download_get('0002605-251120083545085') %>%
    occ_download_import()

# do some QCing
gbif_data_slim <- gbif_data  %>% 
  filter(occurrenceStatus == "PRESENT") %>% 
  dplyr::select(genus, species, verbatimScientificName,
                decimalLatitude, decimalLongitude,
                individualCount, countryCode,stateProvince, 
                year, month, eventDate,
                basisOfRecord)

# correct coords
gbif_points <- st_as_sf(gbif_data_slim, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)

# filter for just cuilicoides parensis
gbif_points_cp <- gbif_points %>% 
  filter(species == "Culicoides paraensis")

```

plot
```{r}
## set up data
# ecologicla niche model suitability from Poongavanan et al 2025
#https://github.com/CERI-KRISP/OROV_Vector_vs_Virus_Modelling/blob/main/Resulting%20Rasters/Americas_Consensus_Map.tif
enm <- terra::rast("data/raw/Americas_Consensus_Map.tif")
enm_df <- as.data.frame(enm, xy = TRUE)
colnames(enm_df) <- c("x", "y", "consensus")

## plot both vector layers
map_vector <- ggplot() +
  geom_raster(data = sdm_df, aes(x = x, y = y, fill = consensus), alpha = .8) +
  geom_sf(data = cs_america, fill = NA, color = alpha("grey40", 0.8), size = 0.35) +
  theme_void(base_size = 13) +
  labs(fill = "Prob.") + 
  theme(legend.title = element_text(face = "bold"),
        plot.margin = margin(0, 0, 0, 0)) +
  scale_fill_paletteer_c("ggthemes::Purple") +
  geom_sf(data = gbif_points_cp,size = 3, alpha = .8) +
  labs(color = "Year", title = "Vector Suitability") +
  coord_sf(xlim = c(-95, -30), ylim = c(-60, 25), expand = FALSE) 


### combine them
map_hostvector_combo <- map_host | map_vector 
```

Make big eco figure
```{r}
map_ecology_combo <- maps_crop_combo/map_hostvector_combo 
```
